---
title: "HMM filtering of scalar SDEs"
author: "Uffe Høgsbro Thygesen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HMM filtering of scalar SDEs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Overview 

We simulate a sample path from the Ito stochastic differential equation governing the abundance of a bacterial population

\[
dX_t = (1- X_t) ~dt + \gamma \sqrt{X_t} ~dB_t
\]

which is the Cox-Ingersoll-Ross model. Then, we simulate random measurements taken at times $0,dt,2~dt, \ldots, T$ so that $Y_i | X_{t_i}$ is Poisson distributed with mean $v `. X_{t_i}$. Then, we pretend that we don't know the states and re-estimate them from these measurements. 

## Simulation of states 

We define the model and simulate the states using the Euler-Marruyama method.

```{r,fig.width=7}
require(SDEtools)

  # Define model. Note abs to handle negative x.
  gamma <- 1
  f = function(x) (1-x)
  g = function(x) gamma * sqrt(abs(x));

  ## Time vector for simulation
  dt = 0.001;
  Tmax = 20;

  tvec = seq(0,Tmax,dt)
  nt <- length(tvec)

  ## Initial condition, chosen somewhat arbitrarily
  x0 <- 0.1

  ## Enforce that the simulated state should be non-negative
  p <- function(x) abs(x)
  
  ## Simulate states 
  sim <- euler(f,g,tvec,x0,p=p)
  
  plot(sim$times,sim$X,type="l",xlab="Time",ylab='Abundance')
```

## Simulation of measurements

Next, we generate some random measurements. Measurements are taken at regular points in time; not at every time point where are state values is simulated. We use a Poisson distribution for the measurements. This correspond to a situation where we count the number of individuals in a small sample from the environment.

```{r,fig.width=7}
## Generate random measurements
  tsample <- 0.1
  vsample <- 1

  sampleIndeces <- round(seq(1,length(tvec),(tsample/dt)))
  tm = tvec[sampleIndeces];
  xtrue <- sim$X[sampleIndeces]
  ymean = vsample * xtrue

  # Generate random measurements
  ymeas <- rpois(length(tm),lambda=ymean)

  plot(sim$times,sim$X,type="l",xlab="Time",ylab='Abundance')
  points(tm,ymeas/vsample,pch='o')
```

## Specification of the likelihood function

To run the filter, we must first specify the likelihood function. We first define this for a general measurement and general state:

```{r,fig.width=7}
  ## State likelihood	
  dl <- function(x,y) dpois(y,lambda=vsample*x)

```

We can now inspect the state likelihood on its own, i.e., without using the process equation. We choose a discretization of state space. We choose to use a grid that has higher resolution near 0, and tabulate the likelihood function.

```{r,fig.width=7}
  ## Choose discretization of state space
  xi <- seq(0,3,0.025)^2                                # Cell Interfaces
  xc <- cell.centers(xi)

  ltab <- outer(xc,ymeas,dl)
  image(tm,xc,t(ltab),xlab="Time",ylab="State")
```

We plot the most likely state, given just the measurements, on top of the true states:

```{r,fig.width=7}
plot(tm,xtrue,type="l")
points(tm,xc[apply(ltab,2,which.max)])
```

There is very little information in the individual observation, because the counts are so low. Therefore, there is a potential gain from estimating the state not just using the measurement taken at the same time, but also the neighbouring measurements. This is particularly so because the measurements are taken quite frequently: The sample interval is 0.1 time unit, while the mean relaxes over a time scale of 1.

## Running the filter

We now use the filter to add the information from the process equation. We use the function HMMfilterSDE, which require a process model as well as a likelihood function. In addition, it needs a distribution of the initial state. We focus on real-time state estimation; i.e., we use only the current and past measurements in the estimation, not future measurements.

```{r,fig.width=7}
  # Advection-diffusion form of the Kolmogorov equations
  D = function(x) 0.5*gamma^2*x;
  u = function(x) f(x) - 0.5*gamma^2;

  ## Specify prior c.d.f. of the initial state, here uniform. Note that it does not have to be normalized.
  phi0 <- function(x)x
  ## Run the filter 
  filter <- HMMfilterSDE(u,D,xi,bc='r',phi0,tm,ymeas,dl)

  ## Show colour plot of posterior c.d.f.
  image(tm,xc,t(apply(filter$psi,1,cumsum)),ylim=c(0,4))
```

## Filtering results

We compare the true states with the estimated ones:

```{r,fig.width=7}
  ## Plot true state at times of measurements
  plot(tm,ymean/vsample,type="l",xlab="t",ylab="x")

  ## Compute and plot posterior mean
  XestMean <- filter$psi %*% xc
  points(tm,XestMean)
```

Notice that the estimated states lag a bit behind the true ones, and that some of the fast fluctuations in the true states do not reflect in the estimated ones. These features are as expected. 

We can compare the expectation and the variance in the posterior distribution:

```{r,fig.width=7}
XestVar <- filter$psi %*% xc^2 - XestMean^2
plot(tm,XestVar)
```

```{r,fig.width=7}
plot(XestMean,XestVar)
```

Note that there seems to be a roughly linear relationship (save the first couple of data points, which reflect transients from the initial condition). This linear relationship reflects both process noise and measurement noise: When the state is large, the process noise is increased, and the variance on the measurements is larger. 

## Perspectives

The function HMMfilterSDE can also gived "smoothed" states, i.e., estimates that are based on all measurements, both past and future. You can compare the performance of the estimates we have shown here, with these smoothed estimates.

You could also estimate the states between the measurements, for example by having a state likelihood function which is constant to 1 whenever no measurement is taken. Then, you could investigate how the posterior variance evolves between times of measurements. 
