---
title: "HMM filtering of scalar SDEs"
author: "Uffe Høgsbro Thygesen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HMM filtering of scalar SDEs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Overview 

We simulate a sample path from the Ito stochastic differential equation governing the abundance of a bacterial population

\[
dX_t = (1- X_t) ~dt + \gamma \sqrt{X_t} ~dB_t
\]

which is the Cox-Ingersoll-Ross model. Then, we simulate random measurements taken at times $0,dt,2~dt, \ldots, T$ so that $Y_i | X_{t_i}$ is Poisson distributed with mean $v `. X_{t_i}$. Then, we pretend that we don't know the states and re-estimate them from these measurements. 

We first simulate states 

```{r,fig.width=7}
require(SDEtools)

  # Define model. Note abs to handle negative x.
  gamma <- 1
  f = function(x) (1-x)
  g = function(x) gamma * sqrt(abs(x));

  # Advection-diffusion form
  D = function(x) 0.5*gamma^2*x;
  u = function(x) f(x) - 0.5*gamma^2;

  ## Time vector for simulation
  dt = 0.001;
  Tmax = 20;

  tvec = seq(0,Tmax,dt)
  nt <- length(tvec)

  ## Initial condition, chosen somewhat arbitrarily
  x0 <- 0.1

  ## Enforce that the simulated state should be non-negative
  p <- function(x) abs(x)
  
  ## Simulate states 
  sim <- euler(f,g,tvec,x0,p=p)
  
  plot(sim$times,sim$X,type="l",xlab="Time",ylab='Abundance')
```

Next, we generate some random measurements: 

```{r,fig.width=7}
## Generate random measurements
  tsample <- 0.1
  vsample <- 1

  sampleIndeces <- round(seq(1,length(tvec),(tsample/dt)))
  tm = tvec[sampleIndeces];
  xtrue <- sim$X[sampleIndeces]
  ymean = vsample * xtrue

  # Generate random measurements
  ymeas <- rpois(length(tm),lambda=ymean)

  plot(sim$times,sim$X,type="l",xlab="Time",ylab='Abundance')
  points(tm,ymeas/vsample,pch='o')
```

We now setup to run the filter. We first specify the likelihood function.

```{r,fig.width=7}
## State likelihood
  dl <- function(x,y) dpois(y,lambda=vsample*x)

  ## Choose discretization of state space
  xi <- seq(0,3,0.025)^2                                # Cell Interfaces
  xc <- head(xi,-1) + 0.5*diff(xi)
```

Let us first inspect the state likelihood on its own, i.e., without using the process equation.

```{r,fig.width=7}
ltab <- outer(xc,ymeas,dl)
image(tm,xc,t(ltab),xlab="Time",ylab="State")
```

We plot the most likely state, given just the measurements, on top of the true states:

```{r,fig.width=7}
plot(tm,xtrue,type="l")
points(tm,xc[apply(ltab,2,which.max)])
```

There is very little information in the individual observation, because the counts are so low. Therefore, there is a potential gain from estimating the state not just using the measurement taken at the same time, but also the neighbouring measurements. This is particularly so because the measurements are taken quite frequently: The sample interval is 0.1 time unit, while the mean relaxes over a time scale of 1.

In the following, we aim to add the information from the process equation. We use the function HMMfilterSDE, which require a process model as well as a likelihood function. In addition, it needs a distribution of the initial state. We focus on real-time state estimation; i.e., we use only the current and past measurements in the estimation, not future measurements.

```{r,fig.width=7}
## Specify prior c.d.f. of the initial state, here uniform. Note that it does not have to be normalized.
  phi0 <- function(x)x
  ## Run the filter 
  filter <- HMMfilterSDE(u,D,xi,bc='r',phi0,tm,ymeas,dl)

  ## Show colour plot of posterior c.d.f.
  image(tm,xc,t(apply(filter$psi,1,cumsum)),ylim=c(0,4))
```

We can compare the true states with the estimated ones.

```{r,fig.width=7}
  ## Plot true state at times of measurements
  plot(tm,ymean/vsample,type="l",xlab="t",ylab="x")

  ## Compute and plot posterior mean
  XestMean <- filter$psi %*% xc
  points(tm,XestMean)
```

Notice that the estimated states lag a bit behind the true ones, and that some of the fast fluctuations in the true states do not reflect in the estimated ones. These features are as expected. 

We can compare the expectation and the variance in the posterior distribution:

```{r,fig.width=7}
XestVar <- filter$psi %*% xc^2 - XestMean^2
plot(XestMean,XestVar)
```

Note that there seems to be a roughly linear relationship. This reflects both process noise and measurement noise: When the state is large, the process noise is increased, and the variance on the measurements is larger. 
