---
title: "SDEtools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SDEtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This [R](https://r-project.org) package provides functionality for stochastic differential equations in terms of: 

* Simulating sample paths
* Computing transition probabilities numerically in 1 and 2 dimensions
* Estimating states (and parameters) for 1D or 2D equations, using Hidden Markov Model methods.
* Solve optimal control problems numerically, using Markov Decision Process machinery.

### Installation

To install the package, open R and run 

```r
devtools::install_github("Uffe-H-Thygesen/SDEtools")
```

Alternatively, clone the github repo to your local machine and "make install" it, if you are familiar with make and makefiles. 

Then load the package with 

```{r setup}
library(SDEtools)
```


## Example

This is a basic example which shows you how to simulate a sample path, and compute the transition probability, for a double well model:

```{r,fig.width=8}
library(SDEtools)
## basic example code
f <- function(x) x-x^3
g <- function(x) 0.5

tv <- seq(0,6,0.01)
sim <- heun(f,g,tv,0)

plot(tv,sim$X,type="l",ylim=c(-1.5,1.5))

xi <- seq(-1.5,1.5,0.01)
xc <- cell.centers(xi)

D <- function(x) 0.5*g(x)^2
G <- fvade(f,D,xi,'r')

rho <- StationaryDistribution(G)

lines(200*rho,xc)
```

## Tour of features

Please see the vignettes which showcase functionality. 
