---
title: "A 2D example: The van der Pol oscillator"
author: "Uffe Høgsbro Thygesen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{A 2D example: The van der Pol oscillator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The following example demonstrates how to simulate solutions to two coupled equations, and how to solve the Kolmogorov equations. 

The case is stochastic van der Pol oscillator: 

\[
dX_t = V_t ~dt , \quad dV_t = [ V_t(\mu-X_t^2) -X_t ]~dt + \sigma ~dB_t
\]

Without noise, this is the standard van der Pol oscillator, which is a classical example of a non-linear system which has a limit cycle. The noise perturbs the oscillations, so that the system will converge to a stochastic steady state. See @Thygesen2023sde for more details about the model.

```{r}
rm(list=ls())
require(SDEtools)

fx <- function(x,v) v
fv <- function(x,v) v*(mu - x^2)-x
gx <- function(x,v) 0
gv <- function(x,v) sigma
```

We set this in the usual state space formalism, defining a state vector $Z_t=(X_t,V_t)$. Then, the drift and noise in this vector equation for $Z_t$ are: 

```{r}
f <- function(xv) c(fx(xv[1],xv[2]),fv(xv[1],xv[2]))
g <- function(xv) c(0,gv(xv[1],xv[2]))
```

We fix the parameters and simulate a sample path:

```{r}
sigma <- 0.5
mu <- 1
T <- 100
dt <- 0.01

times <- seq(0,T,dt)
B <- rBM(times)

xv0 <- c(0,0)

sim <- euler(f,g,times,xv0,B)

plot(times,sim$X[,1],type="l",xlab="Time t",ylab="Position X")
```

We can plot the solution in the phase plane:

```{r}
plot(sim$X[,1],sim$X[,2],type="l",xlab="Position X",ylab="Velocity V")
```

We can also compute the stationary distribution analytically. To this end, we first define a grid.

```{r,fig.width=7,fig.height=7}
xmax <- 1.5*max(abs(sim$X[,1]))
vmax <- 1.5*max(abs(sim$X[,2]))

xgrid <- seq(-xmax,xmax,length=101)
vgrid <- seq(-vmax,vmax,length=91)

xc <- cell.centers(xgrid)
vc <- cell.centers(vgrid)

plot(xmax*c(-1,1),vmax*c(-1,1),type="n",xlab="x",ylab="v")
abline(h=vgrid)
abline(v=xgrid)
```

We next pose the advection/diffusion form of the Kolmogorov equations. Note that the advective field equals the drift, since the noise intensity is state independent. We discretize the equation using the function *fvade2d*, which by default uses reflection at the boundaries.  

```{r}
Dx <- function(x,v) 0
Dv <- function(x,v) 0.5*gv(x,v)^2

G <- fvade2d(fx,fv,Dx,Dv,xgrid,vgrid)
```

We can now compute the stationary distribution. At first, this gives just a vector, so it must be transformed to a field 

```{r}
pi <- unpack.field(StationaryDistribution(G),length(xc),length(vc))

image(xc,vc,t(prob2pdf(pi,xgrid,vgrid)))
```

To compare this with the simulation, it is better to plot the time series as points in the phase plane: 

```{r}
  plot(sim$X[,1],sim$X[,2],pch=".",xlab="x",ylab="v")
```

# References
